Dealing with Stochastic Volatility in Time Series Using the \proglang{R} Package \pkg{stochvol}

### 2014-07-01 inst/doc/article.Rnw
version 0.8-4

A novel and crucial feature of the algorithm implemented in \pkg{stochvol} is the usage of an ``ancillarity-sufficiency interweaving strategy'' (ASIS), which has been brought forward in the general context of state-space models by \cite{yu-men:cen}. ASIS exploits the fact that for certain parameter constellations, sampling efficiency improves substantially when considering a non-centered version of a state-space model.
This fact is commonly known as a reparameterization issue with an entire body of literature attached to it; for an early reference see e.g., \cite{hil-smi:par}.
In the case of the SV model, a move of this kind can be achieved by moving the level $\mupar$ and/or the volatility $\sigmapar$ of log-variance from the state equation (\ref{c2}) to the observation equation (\ref{c1}) through a simple reparameterization of the latent process $\hsvm$. However, in the case of the SV model, it turns out that no single superior parameterization exists. Rather, for some underlying processes, the standard parameterization yields superior results, while for other processes non-centered versions are better. To overcome this issue, the parameter vector $\allpara$ is sampled twice: once in the centered, and once in a noncentered parameterization. This method of ``combining best of different worlds'' allows for efficient inference regardless of the underlying process with one algorithm. For details about the algorithm and empirical results concerning sampling efficiency, see \cite{kas-fru:anc}.

@@

Instead of homoskedastic errors, we now specify the error covariance matrix in Equation~\ref{reg} to be $\Sigmab \equiv \diag{\e^{h_1/2}, \dots, \e^{h_T/2}}$, thus introducing nonlinear dependence between the observations due to the AR(1)-nature of $\hb$. Instead of cooking up an entire new sampler, we adapt the code from above utilizing the \pkg{stochvol} package. To do so, we simply replace the sampling step of $\sigmaepsextrasuper{2}$ from an Inverse-Gamma distribution by a sampling step of $\allpara$ and $\hb$ through a call to \code{.svsample}. This ``dotted'' function is a minimal-overhead version of the regular \code{svsample}. It provides the full sampling functionality of the ``non-dotted'' version, but has slightly different default values, a simplified return value structure and does not perform costly input checks. Thus, it is optimized for repeated calls but needs to be used with proper care. Note that the current draws of the variables need to be passed to the function through \code{startpara} and \code{startlatent}.

@@

>>>>

A novel and crucial feature of the algorithm implemented in \pkg{stochvol} is the usage of an ``ancillarity-sufficiency interweaving strategy'' (ASIS), which has been brought forward in the general context of state-space models by \cite{yu-men:cen}. ASIS exploits the fact that for certain parameter constellations, sampling efficiency improves substantially when considering a non-centered version of a state-space model.
This fact is commonly known as a reparameterization issue with an entire body of literature attached to it; for an early reference see e.g., \cite{hil-smi:par}.
In the case of the SV model, a move of this kind can be achieved by moving the level $\mupar$ and/or the volatility $\sigmapar$ of log-variance from the state equation (\ref{c2}) to the observation equation (\ref{c1}) through a simple reparameterization of the latent process $\hsvm$. However, in the case of the SV model, it turns out that no single superior parameterization exists. Rather, for some underlying processes, the standard parameterization yields superior results, while for other processes non-centered versions are better. To overcome this issue, the parameter vector $\allpara$ is sampled twice: once in the centered, and once in a noncentered parameterization. This method of ``combining best of different worlds'' allows for efficient inference regardless of the underlying process with one algorithm. For more details about the algorithm and empirical results concerning sampling efficiency, see \cite{kas-fru:anc}.

@@

Instead of homoskedastic errors, we now specify the error covariance matrix in Equation~\ref{reg} to be $\Sigmab \equiv \diag{\e^{h_1}, \dots, \e^{h_T}}$, thus introducing nonlinear dependence between the observations due to the AR(1)-nature of $\hb$. Instead of cooking up an entire new sampler, we adapt the code from above utilizing the \pkg{stochvol} package. To do so, we simply replace the sampling step of $\sigmaepsextrasuper{2}$ from an Inverse-Gamma distribution by a sampling step of $\allpara$ and $\hb$ through a call to \code{.svsample}. This ``dotted'' function is a minimal-overhead version of the regular \code{svsample}. It provides the full sampling functionality of the ``non-dotted'' version, but has slightly different default values, a simplified return value structure and does not perform costly input checks. Thus, it is optimized for repeated calls but needs to be used with proper care. Note that the current draws of the variables need to be passed to the function through \code{startpara} and \code{startlatent}.

@@

For an example of using \pkg{stochvol} for univariate SV updates within a factor SV framework, see \cite{kas-etal:ana}.
